---
title: "Final Report"
author: "Mayra Smith-Coronado"
date: "2024-06-26"
output: pdf_document
---
```{r markdown setup, include=FALSE, eval = FALSE}
# to view pdf outputs from the markdown use the following lines of code
# in the console
install.packages(tinytex)
tinytex::install_tinytex()
```

## Intro

The included datasets come from two tables within the DSD relational database: Incarceration and Person. The incarceration table’s unit of analysis is one booking into the jail and includes booking-related data such as the times into and out of the jail. The person table’s unit of analysis is one person and includes demographic information like age and race. The “Person_id” column is the common variable between them. (For this exercise, the real Person_id has been suppressed and replaced with a unique random number to protect identities.)

The point of this exercise is to demonstrate how you think analytically as much as it is to arrive at the “correct” answers. Please provide your best answers to the questions below, using the tools and methods you deem most effective. Please submit written answers in a clear and concise form by the deadline. **Please also share your code so we can review it.**

## Exercise

Consider the year from July 1, 2021 to June 30, 2022 as the analysis period.

```{r project setup, include = FALSE, echo = FALSE}
# markdown settings -------------------------------------------------------
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, include = FALSE, echo = FALSE)

# functions ---------------------------------------------------------------
# create a function that installs and loads packages
install_and_load_packages <- function(packages){
  # find packages missing and install
  new_packages <- setdiff(packages, rownames(installed.packages()))
  install.packages(new_packages)
  # load packages
  invisible(lapply(packages, library, character.only = TRUE))
}


# create a function that replaces or with a new line for graphics
addline_format <- function(x,...){
    gsub(' or ','\n',x)
}

# create a function to simplify donut chart creation
make_donut_charts <- function(data, variable, percent, color = "muted blue",
                              cols = 1, rows = 1, order = "increasing"){
  # errors
  # if(!(color %in% c("purple", "earth green", "muted blue"))) stop("Please select a color that is purple, earth green, or muted blue. For more information, please  refer to the CDPHE style guide.")
  if(nrow(data) > 1 & cols == 1 & rows == 1) stop("There is more than one category in the data frame. Please update either the number of columns and rows to arrange the charts.")
  if(length(order) == 1){
    if(sum(order == "increasing",
           order == "decreasing") == 0) stop("If re-ordering by categories, review categories because they do not match information in the data. Otherwise, set order equal to 'increasing' or 'decreasing'.")
  } else {
    if(!sum(data[, variable] %in% order) == nrow(data)) stop("If re-ordering by categories, review categories because they do not match information in the data. Otherwise, set order equal to 'increasing' or 'decreasing'.")
  }

  # prep and create charts
  if(nrow(data) == 1){
    # gather label information
    emphasis_percent = paste0(data[, percent] * 100, "%")
    emphasis_category = as.character(data[, variable])

    # prep data
    data <- prep_donut_data(data = data, variable_name = variable,
                            percent_name = percent)

    # create plot
    donut_plot <- create_donut(
      data_table = data, variable_name = variable, percent_name = percent,
      color_name = color, emphasis_category = emphasis_category,
      emphasis_percent = emphasis_percent)

    return(donut_plot)
  } else {
    # order data
    data <- order_data(data = data,
                       variable_name = variable,
                       percent_name = percent,
                       order = order)

    plot_list = list()
    # plot_list = NULL
    for(category in 1:nrow(data)){
      # select category data
      temp_data <- data[category, ]

      # gather label information
      emphasis_percent = paste0(temp_data[, percent] * 100, "%")
      emphasis_category = as.character(temp_data[, variable])

      # prep data
      temp_data <- prep_donut_data(data = temp_data, variable_name = variable,
                                   percent_name = percent)

      temp_plot <- create_donut(
        data_table = temp_data, variable_name = variable, percent_name = percent,
        color_name = color, emphasis_category = emphasis_category,
        emphasis_percent = emphasis_percent)

      plot_list[[category]] <- temp_plot
    }

    grid.arrange <- gridExtra::grid.arrange
    do.call("grid.arrange", c(plot_list, ncol = cols, nrow = rows))
  }
}


order_data <- function(data, variable_name, percent_name, order){
  data <- as.data.frame(data)
  if(nrow(data) == 1){
    data[, variable_name] <- as.character(data[, variable_name])
    return(data)
  } else if(length(order) > 1){
    data[, variable_name] <- as.factor(data[, variable_name])
    data[, variable_name] <- factor(data[, variable_name],
                                    levels = order)
    data <- dplyr::arrange(data, !!rlang::sym(variable_name))
    data[, variable_name] <- as.character(data[, variable_name])
    return(data)
  } else if(order == "increasing"){
    data[, variable_name] <- as.character(data[, variable_name])
    data <- data[order(data[, percent_name]), ]
    return(data)
  } else if(order == "decreasing"){
    data[, variable_name] <- as.character(data[, variable_name])
    data <- data[order(data[, percent_name], decreasing = TRUE), ]
    return(data)
  }
}

create_donut <- function(data_table, variable_name, percent_name,
                         color_name, emphasis_category,
                         emphasis_percent){

  data_table %>% ggplot2::ggplot(
    ggplot2::aes(x = 1.5, y = !!dplyr::sym(percent_name),
                 fill = !!dplyr::sym(variable_name))) +
    ggplot2::geom_bar(stat = "identity") +
    # ggplot2::scale_fill_manual(values = covid_emphasis(color_name, "light", TRUE)) +
    ggplot2::scale_fill_manual(values = c("#1F4045", "#EBCDC3")[2:1]) +
    ggplot2::coord_polar(theta="y", start = 0) +
    ggplot2::annotate(geom = 'text', x = 0.5, y = 0,
                      label = emphasis_percent, size = 5) +
    ggplot2::ylab(emphasis_category) +
    theme_minimal() +
    ggplot2::theme(
      # panel parameters
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank(),
      panel.grid.minor.y = ggplot2::element_blank(),
      # caption parameters
      plot.caption = ggplot2::element_text(hjust = 0, size = 8),
      plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
      plot.caption.position =  "plot",
      # title, axis text, and plot text parameters
      title = ggplot2::element_text(size = 14, colour = "#000000"),
      axis.title.x = ggplot2::element_text(face = "bold", size = 14),
      axis.title.y = ggplot2::element_text(face = "bold", size = 14, angle = 0,
                                  vjust = 0.5, hjust = 0),
      axis.text = ggplot2::element_text(size = 10, face = "bold"),
      text = ggplot2::element_text(size = 11, colour = "#000000"),
      # additional parameters
      legend.position = "none",
      axis.line = ggplot2::element_line(
        size = 0.2645833, colour = rgbtohex("166 166 166"))
      # axis.line.x = element_blank(),
      # axis.line.y = element_blank()
    ) +
    ggplot2::theme(
      legend.position = "none",
      axis.line = ggplot2::element_blank(),
      axis.text = ggplot2::element_blank(),
      axis.title.y = ggplot2::element_blank()
    )
}

prep_donut_data <- function(data, variable_name, percent_name){
  # prep data
  data[, variable_name] <- as.character(data[, variable_name])

  # calculate left over percent and re order data
  data <- rbind(data,
                c(paste0("Not ", data[, variable_name]),
                  1 - data[, percent_name])
  )
  data[, percent_name] <- as.numeric(data[, percent_name])
  data <- data[c(2, 1), ]
  data[, variable_name] <- as.factor(data[, variable_name])
  data[, variable_name] <- factor(data[, variable_name], levels = data[, variable_name])

  return(data)
}

rgbtohex <- function(x){
  x <- strsplit(x, " ")[[1]]
  rgb(x[1], x[2], x[3], maxColorValue=255)
}


# load packages -----------------------------------------------------------
project_packages <- c("openxlsx", "janitor", "dplyr", "scales", "lubridate",
                      "ggplot2", "zoo", "survival", "survminer", "coin",
                      "knitr")

install_and_load_packages(project_packages)
```


```{r load data}
# Load in Incarceration Data ----------------------------------------------
# note time is not going to be read in for this data set, only date information
incarceration_data <- read.xlsx("~/osei_data_exercise/01 - data/Incarceration.xlsx") %>% 
  # convert to tibble
  tibble() %>% 
  # update column names to follow snake case naming convention
  janitor::clean_names() %>% 
  # correctly read excel dates as dates instead of numeric values
  mutate(across(.cols = c("date_in", "release_out"),
                .fns = janitor::excel_numeric_to_date))

# check to make sure that there are no duplicate booking numbers and that
# each row represents one booking
incarceration_data %>% 
  count(booking_number) %>% 
  filter(n > 1)

# Load in Person Data -----------------------------------------------------
person_data <- read.xlsx("~/osei_data_exercise/01 - data/Person.xlsx") %>% 
  # convert to tibble
  tibble() %>% 
  # update column names to follow snake case naming convention
  janitor::clean_names() %>% 
  # correctly read excel dates as dates instead of numeric values
  mutate(across(.cols = c("dob"),
                .fns = janitor::excel_numeric_to_date))

# check to make sure that there are no duplicate people and that
# each row represents one person
person_data %>% 
  count(person_id) %>% 
  filter(n > 1)
```


---

### 1. How many total bookings into the jail were there in that time period?

```{r}

bookings_in_analysis_period <- incarceration_data %>% 
  filter(date_in >= "2021-07-01" &
           date_in <= "2022-06-30")

answer1 <- nrow(bookings_in_analysis_period) %>% 
  scales::comma()
```

There were `r answer1` total bookings into the jail from July 1st, 2021 to June 30, 2022.

---

### 2. How many unique people were booked into the jail?

```{r people incarcerated}
answer2 <- bookings_in_analysis_period %>% 
  select(person_id) %>% 
  distinct() %>% 
  count() %>% 
  pull() %>% 
  scales::comma()
```

There were `r answer2` unique people who were booked into the jail from July 1st, 2021 to June 30, 2022.

---

### 3. How many people were in the jail at the moment of the data extraction?

```{r people during extraction}
# gather bookings where a person was booked before the analysis period, but
# has no release date
bookings_still_incarcerated <- incarceration_data %>% 
  filter(date_in < "2021-07-01",
         is.na(release_out))

# gather bookings where a person was booked before the analysis period,
# but they were released within the analysis period
bookings_released_during_extraction <- incarceration_data %>% 
  filter(date_in < "2021-07-01",
         release_out >= "2021-07-01" &
           release_out <= "2022-06-30")

# now create a table with all the bookings that we have identified to 
# have occurred during the extraction and bookings that 
# occurred before the extraction, but were not yet released
bookings_during_extraction <- bookings_in_analysis_period %>% 
  bind_rows(bookings_still_incarcerated) %>% 
  bind_rows(bookings_released_during_extraction)

# now count the number of people in jail at the moment of the 
# data extraction
answer3 <- bookings_during_extraction %>% 
  select(person_id) %>% 
  distinct() %>% 
  count() %>% 
  pull() %>% 
  scales::comma()
```

This would include not only the people who were booked during the analysis period, but the people who were booked before analysis period, but were not yet released. Based on this understanding of the question, I estimated approximately `r answer3` people in the jail at the moment of the data extraction.

---

### 4. Consider the length of stay (LOS): the duration of each booking. Describe the LOS over the year analysis period. What insights (statistical and otherwise) does it provide you about variations in the jail population?

```{r length of stay descriptive stats review}
los_data <- bookings_in_analysis_period %>% 
  # remove bookings with no release date
  filter(!is.na(release_out)) %>% 
  # calculate length of stay
  mutate(length_of_stay = as.numeric(release_out - date_in) + 1)

# median length of stay information
median_los <- median(los_data$length_of_stay)
median_and_less_count <- los_data %>%
  filter(length_of_stay <= median_los) %>% 
  count() %>% 
  pull()
percent_of_bookings_at_median_or_less <- median_and_less_count / nrow(los_data) 

# most frequent length of stay
most_frequent_los <- los_data %>% 
  tabyl(length_of_stay) %>% 
  adorn_pct_formatting() %>% 
  filter(n == max(n))

# third quntile to max los
third_quartile <- quantile(los_data$length_of_stay, probs = 0.75)

los_data %>% 
  mutate(grouped_los = ifelse(length_of_stay >= 10, "10+", length_of_stay),
         grouped_los = factor(grouped_los, levels = c(0:9, "10+"))) %>% 
  tabyl(grouped_los) %>% 
  adorn_pct_formatting()
```

To look at length of stay during the analysis period, I think it is best to look at the bookings that were from July 1, 2021 to June 30, 2022. Bookings missing a release date were excluded from the visual review. Overall, the data seems to suggest that the people who are detained in jail tend to be changing frequently, if a little over half the jail population is only detained for `r median_los` days or less. 

- About `r scales::comma(median_and_less_count)` (`r scales::percent(percent_of_bookings_at_median_or_less, accuracy = 0.1)`) of bookings had a length of stay of at least `r median_los` days.
- The most frequent length of stay being `r most_frequent_los$length_of_stay` days with a total of `r scales::comma(most_frequent_los$n)` (`r most_frequent_los$percent`).
- There are about 25% of bookings that ranged from `r third_quartile` days to `r max(los_data$length_of_stay)`.

```{r include = TRUE}
# examine the distribution of the length of stay for this analysis period
los_data %>% 
  mutate(`LOS <= 3 days` = ifelse(length_of_stay <= 3, "Yes", "No")) %>% 
  ggplot(aes(x = length_of_stay, fill = `LOS <= 3 days`)) +
  geom_histogram(bins = 65) +
  scale_fill_manual(values = c("#1F4045", "#EBCDC3")[2:1],
                    guide = guide_legend(reverse = TRUE)) +
  theme_minimal() +
  scale_y_continuous("Frequency", labels = scales::comma) +
  scale_x_continuous("Length of Stay")  + 
  theme(legend.position = "top")
```
From reviewing the distribution of the length of stay, it was determined that our vairable of interest was positively skewed. To determine what the median length of stay was for this population with 95% certainty, a survival analysis would make the most sense. In this scenario, we would be trying to understand the amount of time it takes a person who was booked to be released from jail during our analysis period.

```{r, length of stay survival analysis}
# create the event flag and create a time flag that represents length of stay
# remembering that this length of stay is until the end of the study period
survival_data <- bookings_in_analysis_period %>% 
  mutate(event = case_when(
    is.na(release_out) ~ 0,
    release_out > "2022-06-30" ~ 0,
    T ~ 1)) %>% 
  mutate(time = case_when(
    is.na(release_out) ~ as.numeric(ymd("2022-06-30") - date_in) + 1,
    release_out > "2022-06-30" ~ as.numeric(ymd("2022-06-30") - date_in) + 1,
    T ~ as.numeric(release_out - date_in) + 1
  )) %>% 
  select(booking_number, time, event)

# check the number of events (releases) that occurred during the analysis period
survival_data %>% 
  tabyl(event) %>% 
  adorn_pct_formatting()

# calculate the Kaplan-Meier estimate
km <- survival::survfit(Surv(time, event) ~ 1,
  data = survival_data
)

los_ci_lower <- unname(quantile(km, probs = 0.5)$lower)
los_ci_upper <- unname(quantile(km, probs = 0.5)$upper)
```

From a survival analysis, the median release time for a person can be estimated to be `r unname(median(km))` days, with a 95% CI [`r los_ci_lower`,`r los_ci_upper`]. One thing to note is that this median does differ from our review of the descriptive statistics because in this data set we kept bookings with a missing date where these bookings were previously excluded. 

```{r include = TRUE}
# examine a graph of the probablity of release
survminer::ggsurvplot(km,
  conf.int = FALSE,
  surv.median.line = "hv",
  legend = "none"
)
```

---

\pagebreak

### 5. What was the average daily population in the jail during that year? Daily population ought to include anyone who spent even one minute in the jail in a given day. Please describe the methods/approach you used to answer this question. What tool did you use? What functions or other capabilities within the tool? (That is, help another analyst replicate what you did. Sharing code with your answers is encouraged but by no means required.)

Use a for loop to create a data set where if a booking occurred from Monday to Friday (5 days) the booking will be spread across 5 rows. This will create a data set where we can look at who was in jail each day of the analysis period. In order to correctly capture all the individuals in jail during this time, we will use the data set `bookings_during_extraction` created to answer question 3. The for loop I used can be found below. 

```{r daily booking dataset, include = TRUE, echo = TRUE}
# initialize table
booking_long = NULL

for(ith_booking in 1:nrow(bookings_during_extraction)){
  
  # gather the current bookings information
  booking = bookings_during_extraction[ith_booking, ]
  
  # if release date is missing, set this to the last day of the analysis period
  # to capture each individual, even people who are still in jail after the analysis time constraint
  if(!is.na(booking$release_out)) {
    release_out <- booking$release_out
  } else {
     release_out = as.Date("2022-06-30")
  }
  
  # convert the booking row into a long table where each row represents a day
  # the person was in jail based on their current booking.
  dates_incarcerated = seq(booking$date_in, release_out, by = "1 day")
  ith_booking_long <- tibble(date = dates_incarcerated,
                             booking_number = booking$booking_number,
                             person_id = booking$person_id)
  
  # combine the current bookings table with the larger data set.
  booking_long <- booking_long %>%
    bind_rows(ith_booking_long)
}
```

Now that we have a data set where a bookings length of stay is spread out by date, we can look at the daily population. We can do this by counting the number of bookings by date. This will provide us the daily population for our analysis period.

```{r daily population, include = TRUE, echo = TRUE}
# count the number of people in jail each day
bookings_by_day <- booking_long %>% 
  filter(date >= "2021-07-01" & date <= "2022-06-30") %>% 
  arrange(date) %>% 
  group_by(date) %>% 
  count() %>% 
  ungroup() %>% 
  rename(daily_population = n)

# What is the average daily population for the analysis period?
(average_daily_population <- mean(bookings_by_day$daily_population))
```

From the calculations above, we find that the average daily population in the jail during our analysis period is `r scales::comma(average_daily_population)`. By creating a data set in this way, we can also examine the trend of the daily population overtime. Examples of some graphics are shown below.

```{r daily population graphics, include = TRUE}
# what does the daily population look like overtime? note using a moving average
# to smooth the estimates a little
bookings_by_day %>% 
  mutate(seven_day_moving_average = zoo::rollmean(daily_population, k = 7, fill = NA)) %>% 
  ggplot(aes(x = date, y = daily_population)) + 
  geom_point(alpha = 10, size = 2, color = "#EBCDC3") +
  geom_line(aes(x = date, y = seven_day_moving_average), color = "#1F4045", size = 1.5) +
  scale_y_continuous("Daily Population", limit = c(1000, 2000), labels = comma) + 
  scale_x_date("Date", date_breaks = "1 month", date_labels = "%b %y") +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank()) +
  labs(title = "Daily population of jail overtime.",
       subtile = "Analysis period is from July 1st, 2021 to June 30th, 2022")

# View the daily population against the average daily population. Around
# what months is the daily population greater than the average?
bookings_by_day %>% 
  ggplot(aes(x = date, y = daily_population)) + 
  geom_hline(aes(yintercept = average_daily_population), color = "#1F4045", size = 1, linetype = "dashed") + 
  geom_point(alpha = 0.5, size = 2, color = "#EBCDC3") +
  geom_text(x = ymd("2022-06-15"), y = 1600, size = 3.5,
            label = paste0("Average: ", scales::comma(average_daily_population)), color = "#1F4045") + 
  scale_y_continuous("Daily Population", limit = c(1000, 2000), labels = comma) + 
  scale_x_date("Date", date_breaks = "1 month", date_labels = "%b %y") +
  theme(panel.grid.minor.x = element_blank()) +
  labs(title = "Daily population of jail vvertime compared to the average population for the year.",
       subtile = "Analysis period is from July 1st, 2021 to June 30th, 2022") +
  theme_minimal() + 
  theme(panel.grid.minor.x = element_blank()) 
```

---

### 6. Which day during that year had the lowest daily population? What was the population that day? Which day had the highest daily population? What was the population that day?

```{r}
# date with the lowest daily population 
lowest_daily_pop <- bookings_by_day %>% 
  filter(daily_population == min(daily_population))

# date with the highest daily  population 
highest_daily_pop <- bookings_by_day %>% 
  filter(daily_population == max(daily_population))
```

`r format(lowest_daily_pop$date, "%B %d, %Y")` was the day with the lowest daily population of `r comma(lowest_daily_pop$daily_population)`. The day with the highest daily population was `r format(highest_daily_pop$date, "%B %d, %Y")` with a population of `r comma(highest_daily_pop$daily_population)`.

---

\pagebreak

### 7. Please provide a basic analysis of jail demographics during that year period. Are there meaningful statistical relationships among the different groups in the jail?

For this analysis, we are assuming that each booking is a new person, where people who have been booked more than once will be double counted.

```{r booking demographic review}
booking_demographics <- bookings_in_analysis_period %>% 
  # remove people with a missing release date
  filter(!is.na(release_out)) %>% 
  # add in demographic data
  left_join(person_data) %>% 
  mutate(age_at_booking = floor(interval(dob, date_in) / years(1))) %>%
  # calculate length of stay
  mutate(length_of_stay = as.numeric(release_out - date_in) + 1) %>% 
  mutate(gender_ordered = factor(gender, levels = c("Male", "Female")))

# descriptive stats by gender & length of stay distribution
gender_descriptives <- booking_demographics %>% 
  group_by(gender_ordered) %>% 
  summarise(n = n(),
            median_los = median(length_of_stay),
            los_iqr = IQR(length_of_stay)) %>% 
  ungroup() %>% 
  mutate(total = sum(n),
         pct_of_total = round(n/total, digits = 2)) %>% 
  select(gender_ordered, n, pct_of_total, median_los, los_iqr) %>% 
  arrange(n)

# descriptive stats by race & length of stay distribution
race_descriptives <- booking_demographics %>% 
  group_by(race) %>% 
  summarise(n = n(),
            median_los = median(length_of_stay),
            los_iqr = IQR(length_of_stay)) %>% 
  ungroup() %>% 
  mutate(total = sum(n),
         pct_of_total = round(n/total, digits = 2)) %>% 
  select(race, n, pct_of_total, median_los, los_iqr) %>% 
  arrange(n)

# descriptive stats by age
age_descriptives <- booking_demographics %>% 
  filter(!is.na(age_at_booking)) %>% 
  summarise(median_age = median(age_at_booking),
            mean_age = mean(age_at_booking)) %>% 
  ungroup() %>% 
  select(median_age, mean_age)
```

- The majority of the jail population was men (`r scales::percent(gender_descriptives$pct_of_total[which(gender_descriptives$gender_ordered == "Male")])`) where their length of stay had a larger variation (IQR = `r gender_descriptives$los_iqr`) compared to women.

```{r gender donuts, include = TRUE, out.height="25%", fig.align = "center"}
gender_descriptives %>% 
  mutate(category = paste0(gender_ordered, "\n ", "n = ", scales::comma(n))) %>% 
  select(category, pct_of_total) %>% 
  make_donut_charts(variable = "category", 
                    percent = "pct_of_total", 
                    rows = 1,
                    cols = 2,
                    order = "decreasing")
```


```{r gender distribution, include = TRUE, out.height="40%", fig.align = "center"}
booking_demographics %>% 
  ggplot(aes(x = length_of_stay, y = gender_ordered)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  scale_x_continuous("Length of Stay") +
  scale_y_discrete("Gender")
```
\pagebreak

- A large proportion of the jail population was white (`r scales::percent(race_descriptives$pct_of_total[which(race_descriptives$race == "White")])`) and the variance of length of stay among the other race's seemed consistent, with the exception of the Asian or Pacific Islander group.

```{r race donuts, include = TRUE, out.height="50%", fig.align = "center"}
race_descriptives %>% 
  mutate(race = addline_format(race), 
         race = ifelse(race == "Asian\nPacific Islander",
                       "Asian\nPacific\nIslander",
                       race), 
         category = paste0(race, "\n ", "n = ", scales::comma(n))) %>% 
  select(category, pct_of_total) %>% 
  make_donut_charts(variable = "category", 
                    percent = "pct_of_total", 
                    rows = 1,
                    cols = 5,
                    order = "decreasing")
```

```{r race distribution, include = TRUE, out.height="40%", fig.align = "center"}
booking_demographics %>% 
  mutate(race = addline_format(race)) %>% 
  ggplot(aes(x = length_of_stay, y = race)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  scale_x_continuous("Length of Stay") +
  scale_y_discrete("Race")
```
- The age of the people in jail during this time period was also positively skewed where the jail population tends to be younger. At least 50% of the population was `r age_descriptives$median_age` years old and younger.

```{r age distribution, include = TRUE, out.height="50%", fig.align = "center"}
booking_demographics %>% 
  mutate(`Age <= 33` = ifelse(age_at_booking <= 33, "Yes", "No")) %>% 
  ggplot(aes(x = age_at_booking, fill = `Age <= 33`)) +
  geom_histogram(bins = 65) +
  scale_fill_manual(values = c("#1F4045", "#EBCDC3")[2:1],
                    guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous("Daily Population", labels = comma) +
  scale_x_continuous("Age at Time of Booking") + 
  theme_minimal() + 
  theme(legend.position = "top")
```

```{r booking median analysis}
# is the median length of stay greater for male detainees than women detainees
women_pvalue <- wilcox.test(formula = length_of_stay ~ gender_ordered,
            data = booking_demographics,
            extact = FALSE)$p.value

# is the median length of stay less for white than other races?

## Asian or Pacific Islander
white_and_asian_pacific_islander <- booking_demographics %>% 
  filter(race == "White" |
           race == "Asian or Pacific Islander") %>% 
  mutate(race_ordered = factor(race, levels = c("White", "Asian or Pacific Islander")))

asian_pacific_islander_pvalue <- wilcox.test(formula = length_of_stay ~ race_ordered,
            data = white_and_asian_pacific_islander,
            extact = FALSE)$p.value

## American Indian or Alaska Native
white_and_american_indian <- booking_demographics %>% 
  filter(race == "White" |
           race == "American Indian or Alaska Native") %>% 
  mutate(race_ordered = factor(race, levels = c("White", "American Indian or Alaska Native")))

american_indian_pvalue <- wilcox.test(formula = length_of_stay ~ race_ordered,
            data = white_and_american_indian,
            extact = FALSE)$p.value

## Unknown 
white_and_unknown <- booking_demographics %>% 
  filter(race == "White" |
           race == "Unknown") %>% 
  mutate(race_ordered = factor(race, levels = c("White", "Unknown")))

unknown_pvalue <- wilcox.test(formula = length_of_stay ~ race_ordered,
            data = white_and_unknown,
            extact = FALSE)$p.value

## Black
white_and_black <- booking_demographics %>% 
  filter(race == "White" |
           race == "Black") %>% 
  mutate(race_ordered = factor(race, levels = c("White", "Black")))

black_pvalue <- wilcox.test(formula = length_of_stay ~ race_ordered,
            data = white_and_black,
            extact = FALSE)$p.value

# gather all the pvalues
pvalue_table <- tibble(
  `Reference Group` = c("Men",
                      "White",
                      "White",
                      "White",
                      "White"),
  `Comparison Group` = c("Women",
                       "Asian or Pacific Islander",
                       "American Indian or Alaska Native",
                       "Black",
                       "Unknown"),
  `P Value` = c(women_pvalue,
             asian_pacific_islander_pvalue,
             american_indian_pvalue,
             black_pvalue,
             unknown_pvalue)) %>% 
  mutate(`P Value` = case_when(
    `P Value` >= 0.01 ~ as.character(round(`P Value`, digits = 2)),
    `P Value` < 0.0001 ~ "<0.0001",
    `P Value` < 0.001 ~ "<0.001",
    `P Value` < 0.01 ~ "<0.01"
  ))

```

Because of how skewed the data is for length of stay among each demographic, I believe that Wilcoxon rank sum test would be the best way to examine the median by demographic and see if there is a difference between groups. To examine the median for race, people who are white are going to be the control group and compared to the other races.

```{r pvalue table, include = TRUE}
kable(pvalue_table)
```

The Wilcoxon test showed that the medians were significantly different (p `r pvalue_table[which(pvalue_table[,"Comparison Group"] == "Women"), "P Value"] %>% pull()`) between men and women in our jail population where the median length of stay for women was `r gender_descriptives$median_los[which(gender_descriptives$gender_ordered == "Female")]` (IRQ = `r gender_descriptives$los_iqr[which(gender_descriptives$gender_ordered == "Female")]`) and the median length of stay among men was `r gender_descriptives$median_los[which(gender_descriptives$gender_ordered == "Male")]` (IQR = `r gender_descriptives$los_iqr[which(gender_descriptives$gender_ordered == "Male")]`). An additional Wilcoxon test showed that the length of stay median was significantly different (p<`r pvalue_table[which(pvalue_table[,"Comparison Group"] == "American Indian or Alaska Native"), "P Value"] %>% pull()`) between white jail detainees who had a median length of stay of `r race_descriptives$median_los[which(race_descriptives$race == "White")]` (IQR = `r race_descriptives$los_iqr[which(race_descriptives$race == "White")]`) and American Indian or Alaska Natives who had a median length of stay of `r race_descriptives$median_los[which(race_descriptives$race == "Wmerican Indian or Alaska Native")]` (IQR = `r race_descriptives$los_iqr[which(race_descriptives$race == "American Indian or Alaska Native")]`).

---

\pagebreak

### 8. What other insights, useful facts, or questions/concerns did you uncover, if any, in the data?

```{r rebooking}
# how many times are people rebooked (booked at least one time)
rebooking <- los_data %>% 
  arrange(person_id, date_in) %>% 
  group_by(person_id) %>% 
  mutate(frequency_booked = 1:n()) %>% 
  ungroup() %>% 
  arrange(person_id, desc(frequency_booked)) %>% 
  distinct(person_id, .keep_all = T) %>% 
  mutate(booked_multiple_times = ifelse(frequency_booked >= 2, "2+", "1")) 

# review how many people have been booked more than once
rebooked_frequency <- rebooking %>% 
  tabyl(booked_multiple_times) %>% 
  adorn_pct_formatting() %>% 
  filter(booked_multiple_times == "2+")
```

- While working with this data, I was curious about what others have noticed when it comes to length of stay among people who are detained. One similarity that I noticed between this data and the peer reviewed papers that I found published was how young people represented the majority of the jail population. In the paper that I found, a relationship the researchers noticed among their cohort was that the length of stay was longer for people who were younger. I would think it would be valuable to investigate this further and see if this relationship is seen in this study population, since in the demographic section it was visible that the jail population tended to be younger.
- An additional research question I found while getting familiar with this topic was the relationship between length of stay and being detained more than once. Among the bookings in this analysis period, `r scales::comma(rebooked_frequency$n)` (`r rebooked_frequency$percent`) of people were detained more than once so I wonder how this impacted an individuals second stay. This could be examined using a survival analysis.
- Due to the large population of bookings with length of stays that were 10 days or longer (25%), I would like to examine the demographic differences between people with a short, medium, and long length of stay. An analysis that looked at these 3 categories might give us more insight to the relationship between demographics and length of stay.
- Lastly, I would like to make a visual that compares the number of people detained by race in our analysis year to the general ppopulation and maybe narrow this down to the state or county the jail is located to see if the population in the jail mirrors or is different to the general public. An example of the graphic I am thinking about can be found here: https://stephanieevergreen.com/proportion-plots/

